Faster generic IND-CCA2 secure KEM using ``encrypt-then-MAC''

### 2024-09-23 etm-transform/manuscript/main.tex
Draft rewrite of the MAC performance section

\section{Implementation}\label{sec:implementation}
Originally known as Kyber \cite{bos2018crystals}\cite{avanzi2019crystals}, ML-KEM is an \texttt{IND-CCA2} secure key encapsulation mechanism standardized in FIPS 203 by NIST. The \texttt{IND-CCA2} security of ML-KEM is achieved in two steps. First, ML-KEM constructs an \texttt{IND-CPA} secure public key encryption scheme $\texttt{K-PKE}(\keygen, \encrypt, \decrypt)$ whose security is based on the conjectured intractability of the module learning with error (MLWE) problems against both classical and quantum adversaries. Then, the $U^{\not\bot}_m$ variant of the Fujisaki-Okamoto transformation \cite{hofheinz2017modular} is used to construct the KEM $\texttt{MLKEM}(\keygen, \encap, \decap)$ by calling $\texttt{K-PKE}.\keygen, \texttt{K-PKE}.\encrypt, \texttt{K-PKE}.\decrypt$ as sub-routines. Because $\texttt{K-PKE}.\encrypt$ includes substantially more arithmetics than $\texttt{K-PKE}.\decrypt$, by using \textit{re-encryption} and \textit{de-randomization}, ML-KEM's decapsulation routine suffers from computational inefficiency.

We implemented the ``encrypt-then-MAC'' KEM construction using $\texttt{K-PKE}$ as the input PKE and compared its performance against ML-KEM under a variety of scenarios. The experimental data showed that while the ``encrypt-then-MAC'' construction adds a small amount of computational overhead to the encapsulation routine and a small increase in ciphertext size when compared with ML-KEM, it boasts enormous runtime savings in the decapsulation routine, which makes it particularly suitable for deployment in constrained environment.

@@

The CMAC is based on the CBC-MAC, where the block chosen is chosen to be AES-256. To compute a CMAC tag, the message is first broke into 128-bit blocks with appropriate padding. Each block is first XOR'd with the previous block's output, then encrypted under AES using the symmetric key. The final output is XOR'd with a sub key derived from the symmetric key, before being encrypted for one last time. A summary of the computation can be found in figure \ref{fig:aes-256-cmac}

@@

% TODO: What is KMAC

% TODO: We benchmarked MAC performance computing tags on various sizes of ML-KEM ciphertexts

\begin{table}[H]
    \centering
    \caption{MAC performance}\label{tbl:mac-performance}
    \begin{tabular}{|c|c|c|c|c|c|c|c|}
        \hline
        \multicolumn{8}{|c|}{ML-KEM-512} \\
        \hline
        \multicolumn{4}{|c|}{Ciphertext size (bytes): }
        & \multicolumn{4}{|c|}{768} \\
        \hline
        \multicolumn{2}{|c|}{Poly1305} 
        & \multicolumn{2}{|c|}{GMAC} 
        & \multicolumn{2}{|c|}{CMAC} 
        & \multicolumn{2}{|c|}{KMAC} \\
        \hline
        Median & 909 
        & Median & 3899 
        & Median & 6291 
        & Median & 6373\\
        \hline
        Average & 2823 
        & Average & 4859 
        & Average & 6373 
        & Average & 7791 \\
        \hline
        \hline
        \multicolumn{8}{|c|}{ML-KEM-768} \\
        \hline
        \multicolumn{4}{|c|}{Ciphertext size (bytes): }
        & \multicolumn{4}{|c|}{1088} \\
        \hline
        \multicolumn{2}{|c|}{Poly1305} 
        & \multicolumn{2}{|c|}{GMAC} 
        & \multicolumn{2}{|c|}{CMAC} 
        & \multicolumn{2}{|c|}{KMAC} \\
        \hline
        Median & 961 
        & Median & 3899 
        & Median & 7305 
        & Median & 9697 \\
        \hline
        Average & 2704 
        & Average & 4827 
        & Average & 7588 
        & Average & 9928 \\
        \hline
        \hline
        \multicolumn{8}{|c|}{ML-KEM-1024} \\
        \hline
        \multicolumn{4}{|c|}{Ciphertext size (bytes): }
        & \multicolumn{4}{|c|}{1568} \\
        \hline
        \multicolumn{2}{|c|}{Poly1305} 
        & \multicolumn{2}{|c|}{GMAC} 
        & \multicolumn{2}{|c|}{CMAC} 
        & \multicolumn{2}{|c|}{KMAC} \\
        \hline
        Median & 1065 
        & Median & 4055 
        & Median & 8735 
        & Median & 11647 \\
        \hline
        Average & 1809 
        & Average & 5026 
        & Average & 8772 
        & Average & 12186 \\
        \hline
    \end{tabular}
\end{table}

>>>>

\section{Implementation}\label{sec:implementation}
ML-KEM is an \texttt{IND-CCA2} secure key encapsulation mechanism standardized by NIST in FIPS 203. The \texttt{IND-CCA2} security of ML-KEM is achieved in two steps. First, ML-KEM constructs an \texttt{IND-CPA} secure public key encryption scheme $\texttt{K-PKE}(\keygen, \encrypt, \decrypt)$ whose security is based on the conjectured intractability of the module learning with error (MLWE) problems against both classical and quantum adversaries. Then, the $U^{\not\bot}_m$ variant of the Fujisaki-Okamoto transformation \cite{hofheinz2017modular} is used to construct the KEM $\texttt{MLKEM}(\keygen, \encap, \decap)$ by calling $\texttt{K-PKE}(\keygen, \encrypt, \decrypt)$ as sub-routines. Because $\texttt{K-PKE}.\encrypt$ includes substantially more arithmetics than $\texttt{K-PKE}.\decrypt$, by using \textit{re-encryption} and \textit{de-randomization}, ML-KEM's decapsulation routine incurs significant computational cost. Also worth noting is the modification to CRYSTALS-Kyber (as presented in its third round submission to NIST's PQC) before it is standardized: Kyber constructed its KEM routines using the $U^{\not\bot}$ variant of the Fujisaki-Okamoto transform, where the shared secret is derived from hashing both the plaintext and the ciphertext; on the other hand, ML-KEM used the $U^{\not\bot}_m$ variant, where the shared secret is derived from hasing the plaintext alone.

We implemented the ``encrypt-then-MAC'' KEM construction using $\texttt{K-PKE}$ as the input PKE and compared its performance against ML-KEM under a variety of scenarios. The experimental data showed that while the ``encrypt-then-MAC'' construction adds a small amount of computational overhead to the encapsulation routine and a small increase in ciphertext size when compared with ML-KEM, it boasts enormous runtime savings in the decapsulation routine, which makes it particularly suitable for deployment in constrained environment

@@

CMAC is based on the CBC-MAC with the block cipher instantiated from AES-256. To compute a CMAC tag, the message is first broke into 128-bit blocks with appropriate padding. Each block is first XOR'd with the previous block's output, then encrypted under AES using the symmetric key. The final output is XOR'd with a sub key derived from the symmetric key, before being encrypted for one last time. A summary of the computation can be found in figure \ref{fig:aes-256-cmac}

@@

KMAC is defined in NIST SP 800-185 to be based on the family of sponge functions with Keccak permutaiton as the underlying function. We chose KMAC-256, which uses Shake256 as the underlying extendable output functions. KMAC allows variable-length key and tag, but we chose the 256 bits for key length and 128 bits for tag size for consistency with other authenticators.

To isolate the performance characteristics of each authenticator in our instantiation of $\texttt{ML-KEM}^+$, we measured the CPU cycles needed for each authenticator to compute a tag on random inputs whose sizes correspond to the ciphertext sizes of \texttt{ML-KEM}. The measurements are summarized in table \ref{tbl:mac-performance}.

\begin{table}[H]
    \centering
    \footnotesize
    \caption{CPU cycles needed to compute tag on various input sizes}\label{tbl:mac-performance}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 768 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 909 & 2823 \\
        \hline
        GMAC & 3899 & 4859 \\
        \hline
        CMAC & 6291 & 6373 \\
        \hline
        KMAC & 6373 & 7791 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1088 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 961 & 2704 \\
        \hline
        GMAC & 3899 & 4827 \\
        \hline
        CMAC & 7305 & 7588 \\
        \hline
        KMAC & 9697 & 9928 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1568 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 1065 & 1809 \\
        \hline
        GMAC & 4055 & 5026 \\
        \hline
        CMAC & 8735 & 8772 \\
        \hline
        KMAC & 11647 & 12186 \\
        \hline
    \end{tabular}
\end{table}
